---
title: "Model Interrogation"
author: Stephen E. Lane
output:
  html_document:
    theme: simplex
    highlight: zenburn
---

<!-- Time-stamp: <2017-04-26 08:06:51 (slane)> -->

```{r,knitr-setup,echo=FALSE,warning=FALSE,message=FALSE,cache=FALSE,results="hide"}
ipak <- function(pkg){
    ## Check for github packages (throw away github username)
    chk.git <- gsub(".*/", "", pkg)    
    new.pkg <- pkg[!(chk.git %in% installed.packages()[, "Package"])]
    if(!(length(new.pkg) == 0)){
        git.ind <- grep("/", new.pkg)
        if(length(git.ind) == 0){
            install.packages(new.pkg, dependencies = TRUE,
	                     repos = "https://cran.csiro.au/")
        } else {
            devtools::install_github(new.pkg[git.ind])
        }
    }
    sapply(chk.git, require, character.only = TRUE)
}
## Add github packages using gitname/reponame format
packages <- c("dplyr", "tidyr", "ggplot2", "rstan")
ipak(packages)
knitr::opts_chunk$set(error = FALSE, warning = FALSE, cache = TRUE,
                      message = FALSE, echo = FALSE)
options(digits = 2, width = 100)

```

## Introduction

In this document, I detail the model interrogation for the *blistering barnacles* (otherwise known as *biofouling*) project. For the introduction and set up of the problem, see the [manuscript](file:../manuscripts/censored-mle.pdf). This document expands upon the modelling description as given in that paper.

## Models

The basic model (which we denote M0) included the location of the measurement as a level one fixed effect, and a modelled intercept for the vessel ID:

$$
\begin{align}
  Y_{i} & \sim \text{log-normal}(\mu_{i}, \sigma) \\
  \mu_{i} & = \mu + \alpha_{j[i]} + \beta^{l1} \cdot \text{location1}_{i} + \beta^{l2} \cdot \text{location2}_{i} \\
  \alpha_{j} & \sim \text{cauchy}(0, \sigma_{\alpha}) \\
  \mu & \sim N(0, 5) \\
  \beta^{l1},\beta^{l2} & \sim t_{3}(0, 1) \\
  \sigma,\sigma_{\alpha} & \sim \text{cauchy}(0, 2.5)
\end{align}
$$

where $\text{location1}_{i},\text{location2}_{i}$ are indicator variables for whether the observation was measured at the keel or the rudder; observations made at the hull are aliased with the intercept, $\mu$. Summary output from this model is given below.

```{r}
m0 <- readRDS("../data/censored-mle-m0-scaled.rds")
print(m0, digits = 2, include = FALSE,
      par = c("alphaBoat", "yCens", "muHat", "muHatCens", "log_lik",
              "alphaHat"))

```

The next model fit (M1) included all vessel-level predictors as main effects in a regression on the vessel-level intercept:

$$
\begin{align}
  \alpha_{j} & \sim \text{cauchy}(\alpha^{*}_{j}, \sigma_{\alpha}) \\
  \alpha^{*}_{j} & = \beta^{d1} \cdot \text{days1}_{i} + \beta^{d2} \cdot  \text{days2}_{i} + \beta^{m} \cdot \text{midTrips}_{i} + \beta^{h} \cdot \text{hullSA}_{i} + \beta^{p1} \cdot \text{paintType1}_{i} + \beta^{p2} \cdot \text{paintType2}_{i} + \\
  & \quad \beta^{t1} \cdot \text{vesselType1}_{i} + \beta^{t2} \cdot \text{vesselType2}_{i}
\end{align}
$$

where $\text{days1}_{i},\text{days2}_{i}$ are the XXX. Summary output from this model is given below.

```{r}
m1 <- readRDS("../data/censored-mle-m1-scaled.rds")
print(m1, digits = 2, include = FALSE,
      par = c("alphaBoat", "yCens", "muHat", "muHatCens", "log_lik",
              "alphaHat"))

```

## Vessel-level intercept interrogation

In this section we look at the estimated vessel-level intercepts ($\alpha_{j}$) in relation to the vessel-level predictors. From the output above, it is hard to gauge the impact of a product term, but the size of the 50% point estimate of the vessel type coefficients suggest that they may be candidates.

The figure below shows the 50% point estimate (and 80% credible intervals) of the vessel-level intercept against the ordinal predictors, coloured by the vessel type. There is a possibility of an interaction between `days1`/`midTrips` and `boatType`.

```{r}
a1 <- extract(m1, "alphaBoat")$alphaBoat
a1Sum <- t(apply(a1, 2, quantile, probs = c(0.1, 0.5, 0.9)))
a1Dat <- tibble(low = a1Sum[,1], mid = a1Sum[,2], high = a1Sum[,3],
                boatID = 1:nrow(a1Sum))
imps <- readRDS("../data/imputations.rds")
set.seed(13)
lvl2 <- imps[[sample(seq_along(imps), 1)]]$lvl2 %>%
    mutate(
        days1 = as.numeric(scale(days1)),
        days2 = as.numeric(scale(days2)),
        midTrips = as.numeric(scale(midTrips)),
        ApproxHullSA = as.numeric(scale(ApproxHullSA))
    )
a1Dat <- left_join(a1Dat, lvl2) %>%
    gather(type, value, -low, -mid, -high, -boatID, -paintType, -boatType)
slopes <- extract(m1, c("mu", "betaDays1", "betaDays2", "betaMidTrips",
                        "betaHullSA"))
slopes <- sapply(slopes, quantile, probs = 0.5)
slopes <- tibble(intercept = slopes[1], slope = slopes[-1],
                 type = unique(a1Dat$type))
pl <- ggplot(a1Dat, aes(x = value, y = mid, ymin = low, ymax = high,
                        colour = boatType)) +
    geom_pointrange(fatten = 1) +
    facet_wrap(~ type) +
    geom_abline(aes(slope = slope, intercept = intercept), data = slopes) +
    theme_bw()
print(pl)

```

The next figure is similar to above, but it shows the points coloured by paint type. This plot is suggestive of an interaction between `midTrips` and `paintType`.

```{r}
pl <- ggplot(a1Dat, aes(x = value, y = mid, ymin = low, ymax = high,
                        colour = paintType)) +
    geom_pointrange(fatten = 1) +
    facet_wrap(~ type) +
    geom_abline(aes(slope = slope, intercept = intercept), data = slopes) +
    theme_bw()
print(pl)

```

So combining these two exploratory data analyses, and the results from M1, a possible sequence of models to try next would be: M2, which removes `hullSA` from M1; and M3 which adds the `days1:boatType`, `midTrips:boatType` and `midTrips:paintType` interactions to M2.

<p>
  <br><br><br>
  All views presented within are the author's only, and do not necessarily represent the views of CEBRA.
  <br><br>
</p>
